\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography

### 2012-11-30 flux.tex
Change serum prior to diffuse normal.

% notation
\usepackage{amssymb}
\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
\newcommand{\mdssd}{\varphi}						% MDS standard deviation
\newcommand{\diffusionsd}{\sigma}					% Diffusion standard deviation
\newcommand{\driftsd}{\vartheta}					% Drift prior standard deviation
\newcommand{\tree}{\tau}							% Phylogeny
\newcommand{\vn}{n}									% Number of viruses
\newcommand{\sn}{k}									% Number of sera

@@

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\mathbf{X},\mathbf{Y}) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a uniform prior over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$.

@@

As presented, virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
However, these distances represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomena is shown in Figure~\ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}, and so we impose a prior in which virus locations are distributed according to a multivariate normal distribution in which the expectation of the first antigenic dimension increases according to date of virus sampling
\begin{equation} 
	X_{i} \sim \mbox{Normal}( \{\beta \, y_i, 0\}, \Sigma )
\end{equation}
for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus and covariance matrix $\Sigma$ is equal to $\driftsd^2$ along diagonal entries and equal to zero along off-diagonal entries.
Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
The parameter $\driftsd$ determines the breadth of the virus population at each point in time.
We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
We assume a prior on virus locations according to the evolutionary process and leave a flat prior on serum locations
\begin{eqnarray}
	(X_1,\ldots,X_n) &\sim& \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd) \nonumber \\
	Y_j &\sim& \mathrm{Uniform}( -\infty, \infty ),
\end{eqnarray}
where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST}.
The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
Metropolis-Hastings proposals scale $\diffusionsd$.
The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
\begin{equation}
	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
	p(\mathbf{X} | \beta, \vartheta) \;
	p(\mathbf{X} | \tree, \diffusionsd) \; 
	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\vartheta,\diffusionsd)}{ p(\mathbf{H}) }.
\end{equation}

>>>>

% notation
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\point}{f_{\scriptscriptstyle \vert}}	% point likelihood
\newcommand{\threshold}{f_{\textstyle \lrcorner}}	% threshold likelihood
\newcommand{\interval}{f_{\sqcup}}					% interval likelihood
\newcommand{\mdssd}{\varphi}						% MDS standard deviation
\newcommand{\diffusionsd}{\sigma}					% Diffusion standard deviation
\newcommand{\driftsd}{\vartheta}					% Drift prior standard deviation
\newcommand{\tree}{\tau}							% Phylogeny
\newcommand{\vn}{n}									% Number of viruses
\newcommand{\sn}{k}									% Number of sera

@@

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\mathbf{X},\mathbf{Y}) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, where $\mu$ is equal to $\{0,0\}$ and $\Sigma$ is equal to 1000 along diagonal entries and equal to zero along off-diagonal entries.

@@

As presented, virus and serum locations $\mathbf{X}$ and $\mathbf{Y}$ accurately characterize antigenic distances between virus strains based on titers from pairwise HI assays.
However, these distances represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomena is shown in Figure~\ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace the diffuse normal prior on virus locations with a prior in which virus locations are distributed according to a multivariate normal distribution in which the expectation of the first antigenic dimension increases according to date of virus sampling
\begin{equation} 
	X_{i} \sim \mbox{Normal}( \{\beta \, y_i, 0\}, \Sigma )
\end{equation}
for $P=2$, where $y_i$ is the difference in time between virus $i$ and the earliest sampled virus and covariance matrix $\Sigma$ is equal to $\driftsd^2$ along diagonal entries and equal to zero along off-diagonal entries.
Thus, this model assumes that the virus population drifts in a line across the antigenic map at rate $\beta$.
The parameter $\driftsd$ determines the breadth of the virus population at each point in time.
We assume that both the prior slope $\beta$ and the prior precision $1/\driftsd^2$ follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
We further introduce Metropolis-Hastings proposals that scale $\beta$ and $\driftsd$.

@@

We simultaneously model antigenic locations and genetic relatedness by assuming that viruses diffuse across the antigenic map according to a Brownian motion process \cite{Lemey10}.
To do this, we incorporate a prior on virus locations according to the evolutionary process
\begin{equation}
	(X_1,\ldots,X_n) \sim \mbox{Evolutionary Brownian Process}(\tree, \diffusionsd),
\end{equation}
where $\tree$ is a phylogeny specifying tree topology and branch lengths and $\diffusionsd$ is the volatility parameter of the Brownian motion,  set to have equal variance in each antigenic dimension and no correlation across dimensions.
Thus, viruses which are genetically similar are induced to have prior locations close to one another on the antigenic map.
The phylogenetic tree $\tree$ is estimated using sequence data for viruses $1,\ldots,\vn$ according to well established methods implemented in the software package BEAST \cite{BEAST}.
The volatility parameter $\diffusionsd$ is estimated jointly from the genetic and antigenic data.
Parameter $\diffusionsd$ is assumed to follow a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$.
Metropolis-Hastings proposals scale $\diffusionsd$.
The probability of observing virus locations $p(\mathbf{X}|\tree,\diffusionsd)$ is determined through analytical integration across internal states following the methods introduced in \cite{Lemey10}.
Thus, under the full model, the posterior probability of observing virus and serum locations given immunological data and phylogeny is factored
\begin{equation}
	p(\mathbf{X},\mathbf{Y} | \mathbf{H},\tree) = \frac{ p(\mathbf{H}|\mathbf{X},\mathbf{Y},\mathbf{S},\mathbf{V}, \mdssd) \; 
	p(\mathbf{X} | \beta, \vartheta) \;
	p(\mathbf{X} | \tree, \diffusionsd) \; 
	p(\mathbf{Y},\mathbf{S},\mathbf{V},\mdssd,\beta,\vartheta,\diffusionsd)}{ p(\mathbf{H}) }.
\end{equation}
